---
title: "multiomics"
author: "Ranxuan Li"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 300
)
```

# Setup

```{r}
library(tidyverse)
library(cowplot)
library(caret)
library(ROCR)
library(vegan)
library(SuperLearner)
library(bartMachine)
library(bartMachineJARs)
library(IntegratedLearner)
library(BayesCOOP)
library(multiview)
library(curatedMetagenomicData)
library(Maaslin2)
library(MMUPHin)
library(MOFA2)
library(jafar)
library(lme4)

set.seed(123)
theme_set(theme_bw())
options(java.parameters = "-Xmx8000m")
```

# Reuse preprocessing from Lab 5

```{r}
base_url <- paste0(
  "https://raw.githubusercontent.com/",
  "himelmallick/IntegratedLearner/master/data/"
)

data_train <- get(load(url(paste0(base_url, "StelzerDOS.RData"))))
data_test  <- get(load(url(paste0(base_url, "StelzerDOS_valid.RData"))))

ndata_train <- data_train
ndata_train$feature_metadata <- ndata_train$feature_metadata %>%
  dplyr::filter(featureType != "Metabolomics")
ndata_train$feature_table <- ndata_train$feature_table[
  rownames(ndata_train$feature_metadata), , drop = FALSE
]

pos_train <- grep("A", colnames(ndata_train$feature_table), ignore.case = TRUE)
ndata_train$feature_table   <- ndata_train$feature_table[, pos_train, drop = FALSE]
ndata_train$sample_metadata <- ndata_train$sample_metadata[pos_train, , drop = FALSE]

pos_test <- grep("G1", colnames(data_test$feature_table))
data_test$feature_table   <- data_test$feature_table[, pos_test, drop = FALSE]
data_test$sample_metadata <- data_test$sample_metadata[pos_test, , drop = FALSE]

gen_datalist_il <- function(dat) {
  views <- unique(dat$feature_metadata$featureType)
  xList <- lapply(views, function(v) {
    feats <- rownames(dat$feature_metadata)[dat$feature_metadata$featureType == v]
    as.data.frame(t(dat$feature_table[feats, , drop = FALSE]))
  })
  names(xList) <- views
  y <- dat$sample_metadata[, 1]
  if (is.data.frame(y)) y <- y[[1]]
  list(xList = xList, y = y)
}

filter_features_il <- function(x, abd_thresh = 0, prev_thresh = 0.1) {
  keep <- colSums(x > abd_thresh) > nrow(x) * prev_thresh
  x <- x[, keep, drop = FALSE]
  nzv <- caret::nearZeroVar(x)
  if (length(nzv) > 0) x <- x[, -nzv, drop = FALSE]
  x
}

reconstruct_data_IL <- function(xList, y) {
  ft <- as.data.frame(do.call(rbind, lapply(xList, function(x) t(as.data.frame(x)))))
  fmeta <- data.frame(
    featureID = rownames(ft),
    featureType = rep(names(xList), sapply(xList, ncol)),
    stringsAsFactors = FALSE
  )
  rownames(fmeta) <- fmeta$featureID
  smeta <- data.frame(
    Y = as.numeric(y),
    subjectID = colnames(ft),
    stringsAsFactors = FALSE
  )
  rownames(smeta) <- smeta$subjectID
  list(feature_table = ft, feature_metadata = fmeta, sample_metadata = smeta)
}

dl_train_il <- gen_datalist_il(ndata_train)
dl_test_il  <- gen_datalist_il(data_test)

xList_train <- lapply(dl_train_il$xList, filter_features_il)
xList_test  <- lapply(dl_test_il$xList, filter_features_il)
y_train <- dl_train_il$y
y_test  <- dl_test_il$y

keep_features <- vector("list", length(xList_train))
for (i in seq_along(xList_train)) {
  keep_features[[i]] <- intersect(colnames(xList_train[[i]]), colnames(xList_test[[i]]))
  xList_train[[i]] <- as.matrix(xList_train[[i]][, keep_features[[i]], drop = FALSE])
  xList_test[[i]]  <- as.matrix(xList_test[[i]][, keep_features[[i]], drop = FALSE])
}

data_train_processed <- reconstruct_data_IL(xList_train, y_train)
data_test_processed  <- reconstruct_data_IL(xList_test,  y_test)

feature_lookup <- bind_rows(lapply(names(xList_train), function(v) {
  tibble(feature = colnames(xList_train[[v]]), view = v)
}))

dl_train_bc <- BayesCOOP:::gen_datalist(ndata_train)
dl_test_bc  <- BayesCOOP:::gen_datalist(data_test)

xList_train_bc <- lapply(dl_train_bc$xList, function(x) BayesCOOP:::filter_features(x, 0, 0.1))
xList_test_bc  <- lapply(dl_test_bc$xList,  function(x) BayesCOOP:::filter_features(x, 0, 0.1))
y_train_bc <- dl_train_bc$y
y_test_bc  <- dl_test_bc$y

keep_features_bc <- vector("list", length(xList_train_bc))
for (i in seq_along(xList_train_bc)) {
  keep_features_bc[[i]] <- intersect(colnames(xList_train_bc[[i]]), colnames(xList_test_bc[[i]]))
  xList_train_bc[[i]] <- as.matrix(xList_train_bc[[i]][, keep_features_bc[[i]], drop = FALSE])
  xList_test_bc[[i]]  <- as.matrix(xList_test_bc[[i]][, keep_features_bc[[i]], drop = FALSE])
}

data_train_bc <- BayesCOOP:::reconstruct_data(xList_train_bc, y_train_bc)
data_test_bc  <- BayesCOOP:::reconstruct_data(xList_test_bc,  y_test_bc)
```

# Problem 1: Vertical Integration of Pregnancy Data

## Baseline IntegratedLearner results from lab

```{r}
fit_IL <- IntegratedLearner(
  PCL_train = data_train_processed,
  PCL_valid = data_test_processed,
  folds = 5,
  seed = 123,
  base_learner = "SL.randomForest",
  base_screener = "All",
  meta_learner = "SL.nnls.auc",
  run_concat = TRUE,
  run_stacked = TRUE,
  verbose = FALSE,
  print_learner = TRUE,
  refit.stack = TRUE,
  family = gaussian()
)

mspe_IL_early <- mean((fit_IL$yhat.test$concatenated - y_test)^2, na.rm = TRUE)
mspe_IL_late  <- mean((fit_IL$yhat.test$stacked - y_test)^2, na.rm = TRUE)
```

## 1(a) Cooperative learning with `multiview`

```{r}
cvfit_mv <- multiview::cv.multiview(
  x_list = xList_train,
  y = y_train,
  family = gaussian(),
  rho = 0.5,
  alpha = 1,
  nfolds = 5,
  type.measure = "mse"
)

best_lambda <- cvfit_mv$lambda.min

fit_mv <- multiview::multiview(
  x_list = xList_train,
  y = y_train,
  family = gaussian(),
  rho = 0.5,
  alpha = 1,
  lambda = best_lambda
)

coef_mv <- as.matrix(multiview:::coef.multiview(fit_mv, s = best_lambda))
beta_vec <- coef_mv[, 1]
intercept_mv <- beta_vec[1]
beta_no_intercept <- beta_vec[-1]
names(beta_no_intercept) <- rownames(coef_mv)[-1]

Xtest_concat <- do.call(cbind, xList_test)
common_feats <- intersect(colnames(Xtest_concat), names(beta_no_intercept))
Xtest_use <- Xtest_concat[, common_feats, drop = FALSE]
beta_use <- beta_no_intercept[common_feats]

yhat_mv <- as.numeric(intercept_mv + as.matrix(Xtest_use) %*% beta_use)
mspe_mv <- mean((yhat_mv - y_test)^2, na.rm = TRUE)

coef_mv_df <- tibble(
  feature = names(beta_no_intercept),
  beta = as.numeric(beta_no_intercept)
) %>%
  filter(beta != 0) %>%
  left_join(feature_lookup, by = "feature") %>%
  arrange(beta)

n_nonzero_mv <- nrow(coef_mv_df)

knitr::kable(
  tibble(Method = "multiview (Intermediate)", MSPE = mspe_mv, Nonzero = n_nonzero_mv),
  digits = 4
)
```

```{r}
ggplot(coef_mv_df, aes(x = reorder(feature, beta), y = beta, fill = view)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "multiview nonzero coefficients",
    x = NULL,
    y = "Coefficient"
  )
```

## 1(b) BayesCOOP with Bayesian bootstrap

```{r}
fit_bc <- BayesCOOP::BayesCOOP(
  data_train = data_train_bc,
  family = "gaussian",
  ss = c(0.05, 1),
  group = TRUE,
  bb = TRUE,
  alpha_dirich = 1,
  bbiters = 110,
  bbburn = 10,
  maxit = 100,
  warning = TRUE,
  verbose = FALSE,
  control = list()
)

beta_hat <- unlist(fit_bc$beta_postmed)
intercept_bc <- if ("(Intercept)" %in% names(beta_hat)) beta_hat["(Intercept)"] else 0
beta_no_intercept <- beta_hat[setdiff(names(beta_hat), "(Intercept)")]

Xtest_bc <- t(as.matrix(data_test_bc$feature_table))
common_feats_bc <- intersect(colnames(Xtest_bc), names(beta_no_intercept))
Xtest_use_bc <- Xtest_bc[, common_feats_bc, drop = FALSE]
beta_use_bc <- beta_no_intercept[common_feats_bc]
Xtest_centered <- scale(Xtest_use_bc, center = TRUE, scale = FALSE)

bc_pred <- as.numeric(intercept_bc + Xtest_centered %*% beta_use_bc)
mspe_bc <- mean((bc_pred - y_test_bc)^2, na.rm = TRUE)

beta_med <- fit_bc$beta_postmed
beta_samp <- fit_bc$beta_samples
if (!is.null(beta_med)) beta_med <- unlist(beta_med)

if (!is.null(beta_med) && !is.null(beta_samp) && (is.matrix(beta_samp) || is.data.frame(beta_samp))) {
  bc_ci <- tibble(
    feature = names(beta_med),
    beta = as.numeric(beta_med),
    lo = apply(beta_samp, 2, quantile, probs = 0.025, na.rm = TRUE),
    hi = apply(beta_samp, 2, quantile, probs = 0.975, na.rm = TRUE)
  )
} else if (!is.null(beta_med)) {
  bc_ci <- tibble(
    feature = names(beta_med),
    beta = as.numeric(beta_med),
    lo = NA_real_,
    hi = NA_real_
  )
} else {
  bc_ci <- tibble(feature = character(), beta = numeric(), lo = numeric(), hi = numeric())
}

bc_coef_df <- bc_ci %>%
  filter(feature != "(Intercept)", beta != 0) %>%
  left_join(feature_lookup, by = "feature") %>%
  arrange(beta)

knitr::kable(
  tibble(Method = "BayesCOOP (bb=TRUE)", MSPE = mspe_bc, Nonzero = nrow(bc_coef_df)),
  digits = 4
)
```

```{r}
if (nrow(bc_coef_df) > 0) {
  ggplot(bc_coef_df, aes(x = reorder(feature, beta), y = beta, fill = view)) +
    geom_col() +
    geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.2) +
    coord_flip() +
    labs(
      title = "BayesCOOP selected coefficients with 95% credible intervals",
      x = NULL,
      y = "Posterior median coefficient"
    )
}
```

BayesCOOP was successfully fitted and produced posterior coefficient estimates. However, because the package applies internal transformations and hierarchical shrinkage, the manually reconstructed out-of-sample predictions were unstable. Therefore, I interpret BayesCOOP mainly from an inference perspective rather than direct predictive performance.

## 1(c) JAFAR

```{r}
scale_with_train <- function(x_train, x_test) {
  mu <- colMeans(x_train)
  sdv <- apply(x_train, 2, sd)
  sdv[sdv == 0] <- 1
  list(
    train = scale(x_train, center = mu, scale = sdv),
    test  = scale(x_test,  center = mu, scale = sdv)
  )
}

scaled_views <- lapply(seq_along(xList_train), function(i) {
  scale_with_train(xList_train[[i]], xList_test[[i]])
})

Xtrain_jafar <- lapply(scaled_views, `[[`, "train")
Xtest_jafar  <- lapply(scaled_views, `[[`, "test")
m <- length(Xtrain_jafar)
y_train_center <- y_train - mean(y_train)

fit_jafar <- gibbs_jafar(
  Xtrain_jafar,
  y = y_train_center,
  K0 = 25,
  K0_m = rep(20, m),
  tMCMC = 10000,
  tBurnIn = 5000,
  tThin = 10
)

yhat_jafar_center <- predict_y(Xtest_jafar, fit_jafar)
pred_center <- colMeans(yhat_jafar_center$mean)
yhat_jafar <- as.numeric(pred_center) + mean(y_train)
mspe_jafar <- mean((yhat_jafar - y_test)^2, na.rm = TRUE)

knitr::kable(
  tibble(Method = "JAFAR", MSPE = mspe_jafar),
  digits = 4
)
```

## 1(d) Summary comparison

```{r}
mspe_tbl_p1 <- tibble(
  Method = c(
    "IntegratedLearner (Late)",
    "IntegratedLearner (Early)",
    "multiview (Intermediate)",
    "JAFAR",
    "BayesCOOP (bb = TRUE)"
  ),
  MSPE = c(mspe_IL_late, mspe_IL_early, mspe_mv, mspe_jafar, NA_real_)
)

knitr::kable(mspe_tbl_p1, digits = 4)
```

```{r}
mspe_tbl_p1_plot <- mspe_tbl_p1 %>%
  filter(is.finite(MSPE), Method != "BayesCOOP (bb = TRUE)")

barplot(
  rev(mspe_tbl_p1_plot$MSPE),
  names.arg = rev(mspe_tbl_p1_plot$Method),
  horiz = TRUE,
  las = 1,
  main = "Problem 1: MSPE comparison",
  xlab = "MSPE"
)
```

Among all methods, the late integration approach achieved the lowest MSPE, indicating the best predictive performance. This suggests that combining information across omics layers at the model level is more effective than early concatenation or intermediate integration for this pregnancy prediction task.

# Problem 2: Meta-analysis of cancer microbiome pathway data

## 2(a) Load pathway abundance and prepare metadata

```{r}
se_pathway <- sampleMetadata |>
  filter(study_name %in% c(
    "FrankelAE_2017", "GopalakrishnanV_2018",
    "LeeKA_2022", "MatsonV_2018",
    "PetersBA_2019", "WindTT_2020"
  )) |>
  returnSamples("pathway_abundance", rownames = "short")

data_meta <- select(as.data.frame(colData(se_pathway)), c("study_name", "ORR", "PFS12", "anti_PD_1"))
data_meta$resvar <- NA

data_meta$anti_PD_1[data_meta$anti_PD_1 == "responder"]     <- "yes"
data_meta$anti_PD_1[data_meta$anti_PD_1 == "non_responder"] <- "no"
data_meta$resvar[!is.na(data_meta$anti_PD_1)] <- data_meta$anti_PD_1[!is.na(data_meta$anti_PD_1)]
data_meta$resvar[!is.na(data_meta$PFS12)]     <- data_meta$PFS12[!is.na(data_meta$PFS12)]
data_meta$resvar[!is.na(data_meta$ORR)]       <- data_meta$ORR[!is.na(data_meta$ORR)]
data_meta <- data_meta[!is.na(data_meta$resvar), c("study_name", "resvar")]
data_meta$study_name <- factor(data_meta$study_name)

data_path <- assay(se_pathway)
data_path <- data_path[, colnames(data_path) %in% rownames(data_meta), drop = FALSE]

data_path <- data_path[
  !grepl("\\|", rownames(data_path)) &
    !grepl("UNMAPPED|UNINTEGRATED", rownames(data_path), ignore.case = TRUE),
  , drop = FALSE
]

data_path <- data_path / 100
```

## 2(b) Batch correction and PERMANOVA

```{r}
fit_adjust_path <- adjust_batch(
  feature_abd = data_path,
  batch = "study_name",
  covariates = "resvar",
  data = data_meta,
  control = list(verbose = FALSE)
)

data_path_adj <- fit_adjust_path$feature_abd_adj

D_before <- vegdist(t(data_path))
D_after  <- vegdist(t(data_path_adj))

set.seed(1)
fit_adonis_before <- adonis2(D_before ~ study_name, data = data_meta)
fit_adonis_after  <- adonis2(D_after  ~ study_name, data = data_meta)

R2_before <- fit_adonis_before$R2[1]
R2_after  <- fit_adonis_after$R2[1]

knitr::kable(
  tibble(
    Stage = c("Before adjustment", "After adjustment"),
    R2 = c(R2_before, R2_after)
  ),
  digits = 4
)
```

```{r}
pcoa_before <- cmdscale(D_before, eig = TRUE)
pcoa_after  <- cmdscale(D_after, eig = TRUE)

ord_before <- as.data.frame(pcoa_before$points)
ord_after  <- as.data.frame(pcoa_after$points)
colnames(ord_before) <- c("PC1", "PC2")
colnames(ord_after)  <- c("PC1", "PC2")
ord_before$Study <- data_meta$study_name
ord_after$Study  <- data_meta$study_name

pct_before <- round(pcoa_before$eig / sum(pcoa_before$eig) * 100, 1)[1:2]
pct_after  <- round(pcoa_after$eig  / sum(pcoa_after$eig)  * 100, 1)[1:2]

p_before <- ggplot(ord_before, aes(PC1, PC2, color = Study)) +
  geom_point(size = 3) +
  labs(
    title = paste0("Before adjustment (PERMANOVA R2 = ", round(R2_before, 3), ")"),
    x = paste0("Axis 1 (", pct_before[1], "%)"),
    y = paste0("Axis 2 (", pct_before[2], "%)")
  ) +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")

p_after <- ggplot(ord_after, aes(PC1, PC2, color = Study)) +
  geom_point(size = 3) +
  labs(
    title = paste0("After adjustment (PERMANOVA R2 = ", round(R2_after, 3), ")"),
    x = paste0("Axis 1 (", pct_after[1], "%)"),
    y = paste0("Axis 2 (", pct_after[2], "%)")
  ) +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none")

plot_grid(p_before, p_after, ncol = 2)
```

The PCoA plots and PERMANOVA results quantify the extent of study-specific batch effects before and after adjustment. A lower post-adjustment R2 indicates that MMUPHin reduced unwanted study-level variation while preserving the biological signal of interest.

## 2(c) Meta-analytic differential abundance testing

```{r}
fit_meta_path <- lm_meta(
  feature_abd = data_path_adj,
  batch = "study_name",
  data = data_meta,
  exposure = "resvar",
  control = list(
    analysis_method = "CPLM",
    transform = "NONE"
  )
)
#install.packages(c("cli", "tibble", "pillar", "dplyr", "ggplot2"))
#library(cli)

#remove.packages("cli")
#install.packages("cli")
#install.packages(c("tibble", "pillar", "dplyr", "ggplot2"))

res_meta_path <- fit_meta_path$meta_fits

#sig_path <- as_tibble(res_meta_path) %>%
#  filter(qval.fdr < 0.05) %>%
#  arrange(desc(abs(coef)))
#
#knitr::kable(head(sig_path, 20), digits = 4)

nom_path <- res_meta_path %>%
  filter(pval < 0.05) %>%
  arrange(desc(abs(coef)))

knitr::kable(head(nom_path, 20), digits = 4)
```

```{r}
plot_path <- nom_path %>%
  slice_max(order_by = abs(coef), n = 20)

ggplot(plot_path, aes(x = reorder(feature, coef), y = coef)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Significant pathways from MMUPHin meta-analysis",
    x = NULL,
    y = "Meta-analytic coefficient"
  )
```



# Problem 3: Diagonal Integration of iHMP Data (IBD vs. Non-IBD)

## 3(a) Load iHMP data and create binary outcome

```{r p3a-load}
se_species <- sampleMetadata |>
  dplyr::filter(study_name == "HMP_2019_ibdmdb") |>
  returnSamples("relative_abundance", rownames = "short")

se_pathway <- sampleMetadata |>
  dplyr::filter(study_name == "HMP_2019_ibdmdb") |>
  returnSamples("pathway_abundance", rownames = "short")

meta_sp <- as.data.frame(colData(se_species))
meta_pw <- as.data.frame(colData(se_pathway))

common_samples <- intersect(rownames(meta_sp), rownames(meta_pw))
meta <- meta_sp[common_samples, , drop = FALSE]

subject_col <- intersect(c("subject_id", "subjectID"), colnames(meta))[1]
if (is.na(subject_col)) stop("No subject ID column found.")

meta <- meta |>
  dplyr::mutate(
    disease_bin = ifelse(disease %in% c("CD", "UC"), "IBD", "nonIBD"),
    disease_bin = factor(disease_bin, levels = c("nonIBD", "IBD"))
  )

sp_mat <- assay(se_species)[, common_samples, drop = FALSE] / 100

pw_mat <- assay(se_pathway)[, common_samples, drop = FALSE] / 100
pw_mat <- pw_mat[
  !grepl("\\|", rownames(pw_mat)) &
  !grepl("UNMAPPED|UNINTEGRATED", rownames(pw_mat), ignore.case = TRUE),
  , drop = FALSE
]

filter_features <- function(mat, prev_thresh = 0.1) {
  x    <- t(mat)
  keep <- colMeans(x > 0) > prev_thresh
  x    <- x[, keep, drop = FALSE]
  nzv  <- caret::nearZeroVar(x)
  if (length(nzv) > 0) x <- x[, -nzv, drop = FALSE]
  as.matrix(x)
}

sp_x <- filter_features(sp_mat)
pw_x <- filter_features(pw_mat)

meta_use    <- meta[rownames(sp_x), , drop = FALSE]
meta_pw_use <- meta[rownames(pw_x), , drop = FALSE]

knitr::kable(
  tibble(
    Layer      = c("Species", "Pathways"),
    n_samples  = c(nrow(sp_x), nrow(pw_x)),
    n_features = c(ncol(sp_x), ncol(pw_x))
  ),
  caption = "Feature dimensions after filtering"
)
```

## 3(b) Run MaAsLin2 and extract subject-level random effects

```{r p3b-maaslin, results="hide"}
dir.create("maaslin_sp_p3", showWarnings = FALSE)
dir.create("maaslin_pw_p3", showWarnings = FALSE)

fit_sp <- Maaslin2(
  input_data      = sp_x,
  input_metadata  = meta_use,
  output          = "maaslin_sp_p3",
  normalization   = "NONE",
  transform       = "LOG",
  fixed_effects   = "disease_bin",
  random_effects  = subject_col,
  analysis_method = "LM"
)

fit_pw <- Maaslin2(
  input_data      = pw_x,
  input_metadata  = meta_pw_use,
  output          = "maaslin_pw_p3",
  normalization   = "NONE",
  transform       = "LOG",
  fixed_effects   = "disease_bin",
  random_effects  = subject_col,
  analysis_method = "LM"
)
```

```{r p3b-extract-re}

extract_RE_intercept <- function(X, meta, subj_col) {
  meta <- meta[rownames(X), , drop = FALSE]
  form <- as.formula(paste0("y ~ 1 + (1|", subj_col, ")"))

  re_list <- lapply(colnames(X), function(feat) {
    dat   <- meta
    dat$y <- log(X[, feat] + 1e-5)
    if (var(dat$y) < 1e-10) return(NULL)

    mod <- tryCatch(
      suppressWarnings(
        lme4::lmer(form, data = dat, REML = TRUE,
                   control = lme4::lmerControl(optimizer = "bobyqa"))
      ),
      error = function(e) NULL
    )
    if (is.null(mod)) return(NULL)

    re <- tryCatch(lme4::ranef(mod)[[subj_col]], error = function(e) NULL)
    if (is.null(re) || nrow(re) == 0) return(NULL)

    tibble(
      subject = rownames(re),
      feature = feat,
      value   = as.numeric(re[, "(Intercept)"])
    )
  })

  out <- dplyr::bind_rows(re_list)
  if (nrow(out) == 0) stop("No random effects extracted.")

  out |>
    tidyr::pivot_wider(names_from  = feature,
                       values_from = value,
                       values_fill = 0) |>
    tibble::column_to_rownames("subject") |>
    as.matrix()
}

sp_RE <- extract_RE_intercept(sp_x, meta_use,    subject_col)
pw_RE <- extract_RE_intercept(pw_x, meta_pw_use,  subject_col)

## Drop zero-variance columns
sp_RE <- sp_RE[, apply(sp_RE, 2, var) > 0, drop = FALSE]
pw_RE <- pw_RE[, apply(pw_RE, 2, var) > 0, drop = FALSE]

## Subject-level outcome (one label per subject)
y_subj_df <- meta_use |>
  dplyr::select(subject = all_of(subject_col), disease_bin) |>
  dplyr::distinct(subject, .keep_all = TRUE) |>
  dplyr::arrange(subject)

common_subj <- Reduce(intersect,
                      list(rownames(sp_RE), rownames(pw_RE),
                           y_subj_df$subject))

sp_RE     <- sp_RE[common_subj, , drop = FALSE]
pw_RE     <- pw_RE[common_subj, , drop = FALSE]
y_subj_df <- y_subj_df |>
  dplyr::filter(subject %in% common_subj) |>
  dplyr::arrange(subject)

y_subj <- ifelse(y_subj_df$disease_bin == "IBD", 1, 0)
names(y_subj) <- y_subj_df$subject

table(y_subj)

set.seed(123)
folds <- caret::createFolds(
  factor(y_subj, levels = c(0, 1), labels = c("nonIBD", "IBD")),
  k = 5,
  list = TRUE,
  returnTrain = FALSE
)


knitr::kable(
  tibble(
    View       = c("Species RE", "Pathway RE"),
    n_subjects = c(nrow(sp_RE), nrow(pw_RE)),
    n_features = c(ncol(sp_RE), ncol(pw_RE))
  ),
  caption = "Subject-level random effect matrix dimensions"
)
```

## 3(c) IntegratedLearner on subject-level random effects (5-fold CV)

```{r p3-setup}
#
#
```





---

# Problem 4: MOFA+ Factors to Supervised Prediction

## Setup: preprocess pregnancy data

```{r p4-setup}

base_url <- paste0("https://raw.githubusercontent.com/",
                   "himelmallick/IntegratedLearner/master/data/")

data_train_raw <- get(load(url(paste0(base_url, "StelzerDOS.RData"))))
data_test_raw  <- get(load(url(paste0(base_url, "StelzerDOS_valid.RData"))))

data_train_raw$feature_metadata <- data_train_raw$feature_metadata |>
  dplyr::filter(featureType != "Metabolomics")
data_train_raw$feature_table <- data_train_raw$feature_table[
  rownames(data_train_raw$feature_metadata), , drop = FALSE]

pos_tr <- grep("A",  colnames(data_train_raw$feature_table), ignore.case = TRUE)
data_train_raw$feature_table   <- data_train_raw$feature_table[, pos_tr, drop = FALSE]
data_train_raw$sample_metadata <- data_train_raw$sample_metadata[pos_tr, , drop = FALSE]

pos_te <- grep("G1", colnames(data_test_raw$feature_table))
data_test_raw$feature_table   <- data_test_raw$feature_table[, pos_te, drop = FALSE]
data_test_raw$sample_metadata <- data_test_raw$sample_metadata[pos_te, , drop = FALSE]

gen_datalist <- function(dat) {
  views <- unique(dat$feature_metadata$featureType)
  xList <- lapply(views, function(v) {
    feats <- rownames(dat$feature_metadata)[dat$feature_metadata$featureType == v]
    as.data.frame(t(dat$feature_table[feats, , drop = FALSE]))
  })
  names(xList) <- views
  y <- dat$sample_metadata[, 1]
  if (is.data.frame(y)) y <- y[[1]]
  list(xList = xList, y = y)
}

filter_feats <- function(x, abd = 0, prev = 0.1) {
  keep <- colSums(x > abd) > nrow(x) * prev
  x    <- x[, keep, drop = FALSE]
  nzv  <- caret::nearZeroVar(x)
  if (length(nzv) > 0) x <- x[, -nzv, drop = FALSE]
  x
}

reconstruct_IL <- function(xList, y) {
  ft    <- as.data.frame(do.call(rbind,
             lapply(xList, function(x) t(as.data.frame(x)))))
  fmeta <- data.frame(featureID   = rownames(ft),
                      featureType = rep(names(xList), sapply(xList, ncol)),
                      stringsAsFactors = FALSE)
  rownames(fmeta) <- fmeta$featureID
  smeta <- data.frame(Y = as.numeric(y), subjectID = colnames(ft),
                      stringsAsFactors = FALSE)
  rownames(smeta) <- smeta$subjectID
  list(feature_table = ft, feature_metadata = fmeta, sample_metadata = smeta)
}

dl_tr <- gen_datalist(data_train_raw)
dl_te <- gen_datalist(data_test_raw)
xList_tr <- lapply(dl_tr$xList, filter_feats)
xList_te <- lapply(dl_te$xList, filter_feats)
y_tr <- dl_tr$y
y_te <- dl_te$y

for (i in seq_along(xList_tr)) {
  cf <- intersect(colnames(xList_tr[[i]]), colnames(xList_te[[i]]))
  xList_tr[[i]] <- as.matrix(xList_tr[[i]][, cf, drop = FALSE])
  xList_te[[i]] <- as.matrix(xList_te[[i]][, cf, drop = FALSE])
}

data_tr_proc <- reconstruct_IL(xList_tr, y_tr)
data_te_proc <- reconstruct_IL(xList_te, y_te)

cat("Train: features =", nrow(data_tr_proc$feature_table),
    "| subjects =", ncol(data_tr_proc$feature_table), "\n")
cat("Test : features =", nrow(data_te_proc$feature_table),
    "| subjects =", ncol(data_te_proc$feature_table), "\n")
```

## 4(a) Train MOFA+ on pregnancy training data

```{r p4a-mofa}
mofa_input <- list(CYTOF      = t(xList_tr$CYTOF),
                   Proteomics = t(xList_tr$Proteomics))

mofa_obj   <- MOFA2::create_mofa(mofa_input)
data_opts  <- MOFA2::get_default_data_options(mofa_obj)
model_opts <- MOFA2::get_default_model_options(mofa_obj)
train_opts <- MOFA2::get_default_training_options(mofa_obj)

model_opts$num_factors      <- 10
train_opts$seed             <- 123
train_opts$convergence_mode <- "slow"
train_opts$verbose          <- FALSE

mofa_obj <- MOFA2::prepare_mofa(object = mofa_obj,
                                data_options     = data_opts,
                                model_options    = model_opts,
                                training_options = train_opts)

mofa_fit <- MOFA2::run_mofa(mofa_obj, use_basilisk = TRUE)

factors_train <- MOFA2::get_factors(mofa_fit, factors = "all")[[1]]
cat("MOFA factor matrix (train): subjects =", nrow(factors_train),
    "| factors =", ncol(factors_train), "\n")

MOFA2::plot_variance_explained(mofa_fit, plot_total = TRUE)
```

## 4(b) Project train and test into factor space per view

```{r p4b-project}
weights  <- MOFA2::get_weights(mofa_fit,
                               views = c("CYTOF", "Proteomics"),
                               factors = "all")
W_cytof  <- weights$CYTOF
W_prot   <- weights$Proteomics

Ztrain_cytof <- xList_tr$CYTOF      %*% W_cytof
Ztrain_prot  <- xList_tr$Proteomics %*% W_prot
Ztest_cytof  <- xList_te$CYTOF      %*% W_cytof
Ztest_prot   <- xList_te$Proteomics %*% W_prot

knitr::kable(
  tibble(View     = c("CYTOF train","Proteomics train","CYTOF test","Proteomics test"),
         Subjects = c(nrow(Ztrain_cytof),nrow(Ztrain_prot),
                      nrow(Ztest_cytof), nrow(Ztest_prot)),
         Factors  = c(ncol(Ztrain_cytof),ncol(Ztrain_prot),
                      ncol(Ztest_cytof), ncol(Ztest_prot))),
  caption = "Dimension of MOFA factor-score matrices"
)
```

## 4(c) IntegratedLearner on MOFA factor scores

```{r p4c-IL-mofa}
data_mofa_tr <- reconstruct_IL(list(CYTOF = Ztrain_cytof, Proteomics = Ztrain_prot), y_tr)
data_mofa_te <- reconstruct_IL(list(CYTOF = Ztest_cytof,  Proteomics = Ztest_prot),  y_te)

fit_IL_mofa <- IntegratedLearner(
  PCL_train     = data_mofa_tr,  PCL_valid     = data_mofa_te,
  folds         = 5,             seed          = 123,
  base_learner  = "SL.randomForest",
  base_screener = "All",         meta_learner  = "SL.nnls.auc",
  run_concat    = TRUE,          run_stacked   = TRUE,
  verbose       = FALSE,         print_learner = FALSE,
  refit.stack   = TRUE,          family        = gaussian()
)

mspe_mofa_early <- mean((fit_IL_mofa$yhat.test$concatenated - y_te)^2, na.rm = TRUE)
mspe_mofa_late  <- mean((fit_IL_mofa$yhat.test$stacked       - y_te)^2, na.rm = TRUE)

knitr::kable(
  tibble(Method = c("MOFA + IL (Early fusion)", "MOFA + IL (Late fusion)"),
         MSPE   = round(c(mspe_mofa_early, mspe_mofa_late), 4)),
  caption = "Test-set MSPE for MOFA-based IntegratedLearner"
)
```

## 4(d) Full MSPE comparison

```{r p4d-comparison}
fit_IL_raw <- IntegratedLearner(
  PCL_train     = data_tr_proc,  PCL_valid     = data_te_proc,
  folds         = 5,             seed          = 123,
  base_learner  = "SL.randomForest",
  base_screener = "All",         meta_learner  = "SL.nnls.auc",
  run_concat    = TRUE,          run_stacked   = TRUE,
  verbose       = FALSE,         print_learner = FALSE,
  refit.stack   = TRUE,          family        = gaussian()
)

mspe_IL_early_raw <- mean((fit_IL_raw$yhat.test$concatenated - y_te)^2, na.rm = TRUE)
mspe_IL_late_raw  <- mean((fit_IL_raw$yhat.test$stacked       - y_te)^2, na.rm = TRUE)

mspe_all <- tibble(
  Method = c(
    "IntegratedLearner – Early (raw)",
    "IntegratedLearner – Late (raw)",
    "multiview – Intermediate",     # replace NA with mspe_mv from P1(a)
    "JAFAR",                        # replace NA with mspe_jafar from P1(c)
    "MOFA + IL – Early",
    "MOFA + IL – Late"
  ),
  MSPE = c(mspe_IL_early_raw, mspe_IL_late_raw,
           NA_real_, NA_real_,
           mspe_mofa_early, mspe_mofa_late)
) |> dplyr::arrange(MSPE)

knitr::kable(mspe_all, digits = 4,
             caption = "Full MSPE comparison: raw features vs. MOFA factors")
```

```{r p4d-barplot}
plot_df <- mspe_all |>
  dplyr::filter(is.finite(MSPE)) |>
  dplyr::mutate(Method = factor(Method, levels = rev(Method)),
                Type   = ifelse(grepl("MOFA", Method),
                                "MOFA factors", "Raw features"))

ggplot(plot_df, aes(x = MSPE, y = Method, fill = Type)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c("Raw features" = "steelblue",
                               "MOFA factors" = "tomato")) +
  labs(title = "Problem 4 – MSPE: raw features vs. MOFA factor scores",
       x = "MSPE (lower is better)", y = NULL, fill = NULL) +
  theme(legend.position = "bottom")
```